---
title: "US Generation"
author: 
  - "Jamie Kim"
date: "04/10/2023"
output: html_document 
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(message = FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(RTLedu)
library(tidyverse)
library(lubridate)
library(patchwork)
library(kableExtra)
library(gt)
```

```{r Data Frame Collection, echo=FALSE, warning=FALSE}
df <- RTL::eia2tidy_all(tickers = tibble::tribble(~ticker, ~name,
                                                   "TOTAL.ELETPUS.M", "Total_Generation",
                                                   "TOTAL.CLETPUS.M", "Coal",
                                                   "TOTAL.HVETPUS.M", "Hydroelectric",
                                                   "TOTAL.NGETPUS.M", "Natural_Gas",
                                                   "TOTAL.NUETPUS.M", "Nuclear",
                                                   "TOTAL.SOETPUS.M", "Solar",
                                                   "TOTAL.WYETPUS.M", "Wind"),
                                                   key = "b4Of9y6cGeI42KfR3Ydl8seplJ0YqMtYW9QydqMk",
                                                   long = FALSE) %>% 
  dplyr::filter(date >= as.Date("2009-01-01"))
```

```{r Data Frame Manipulation, echo=FALSE, warning=FALSE}
usGen <- df %>% 
  dplyr::arrange(date) %>% 
  dplyr::mutate(Year = year(date)) %>% 
  dplyr::group_by(Year) %>% 
  dplyr::summarise(Yearly_Generation = sum(Total_Generation)) %>% 
  dplyr::filter(Year <= 2022)

long <- pivot_longer(data = df,
                    cols = c(Total_Generation, Coal, Hydroelectric, Natural_Gas, Nuclear, Solar, Wind),
                    names_to = "Sources",
                    values_to = "Values")

```

<div style="text-align: center;">
## **Summary**
</div>

- Electricity generation has slightly grown from 2009 - 2023.
- Electricity generation from coal has significantly decreased while natural gas, solar, and wind have been replacing it
- The share and growth of wind and solar generation has drastically increased
- Total generation has 4 pivotal moments when the trend shifts. Solar generation peaks around summer and wind generation climbs around fall and peaks during the spring.

***
<div style="text-align: center;">
### **Electricity Generation Growth**
</div>

Electricity generation has experienced slow growth, approximately 7.4%, over the past 14 years. Yearly electricity generation was considered to account for and smooth out the seasonality in electricity generation. Since there isn't a full year of data for 2023, it was taken out to ensure consistency.

<div style="text-align: center;">
```{r Is_electricity_generation_growing_or_decreasing, echo=FALSE, message=FALSE, warning=FALSE}

electricity <- usGen %>%
  dplyr::select(Year, Yearly_Generation) %>% 
  ggplot(data = usGen,
         mapping = aes(x = Year,
                       y = Yearly_Generation)) +
  geom_line() +
  stat_smooth(method = "lm", se = FALSE) +
  labs(title = "Yearly Electricity Generation",
       y = "Electricity Generated (Thousand Mwh)")
  
electricity
```
</div>

***
<div style="text-align: center;">
### **Coal Generation and the Sources Replacing it**
</div>

As the US implements climate policies and transitions towards less carbon-intensive electricity generation,
the use of coal as a source of power generation has experienced a rapid decline from 2009 to 2023.

<div style="text-align: center;">
```{r Coal_Graph, echo=FALSE, message=FALSE, warning=FALSE}
p1 <- long %>% 
  dplyr::filter(Sources == "Coal") %>% 
  ggplot(mapping = aes(x = date,
                       y = Values)) +
  geom_line() +
  scale_x_date(date_labels = "%Y", date_breaks = "3 years") +
  labs(title = "Electricity Generated from Coal",
       y = "Electricity Generated (Thousand Mwh)",
       x = "Year") +
  geom_smooth(method = "lm", se = FALSE)
p1
```
</div>

As a result of these policies, the US has increased its reliance on Natural Gas, Solar, and Wind to generate their electricity. Although natural gas is a fossil fuel, it emits fewer carbon emissions compared to coal.

<div style="text-align: center;">
```{r Sources_Replacing_Coal, echo=FALSE, message=FALSE, warning=FALSE}
p.wrap <- long %>%
  dplyr::filter(Sources %in% c("Solar", "Wind", "Natural_Gas", "date")) %>% 
  ggplot(aes(x = date, 
             y = Values,
             col = Sources)) +
  geom_line() +
  facet_wrap(Sources ~ .,
             ncol = 1,
             scales = "free") +
  scale_x_date(date_labels = "%Y", date_breaks = "3 years") +
  geom_smooth(method = "lm", se = FALSE, size = 0.25) +
  labs(title = "Sources Replacing Coal",
       y = "Electricity Generated (Thousand Mwh)",
       x = "Year")
p.wrap
```
</div>

On the contrary, Hydroelectricity and Nuclear energy have stayed relatively stable as a source of electricity.

<div style="text-align: center;">
```{r Hydroelectricity_and_Nuclear_Graph, echo=FALSE, message=FALSE, warning=FALSE}
p.wrap2 <- long %>%
  dplyr::filter(Sources %in% c("Nuclear", "Hydroelectric", "date")) %>% 
  ggplot(aes(x = date, 
             y = Values,
             col = Sources)) +
  geom_line() +
  facet_wrap(Sources ~ .,
             ncol = 1,
             scales = "free") +
  scale_x_date(date_labels = "%Y", date_breaks = "3 years") +
  geom_smooth(method = "lm", se = FALSE, size = 0.25) +
  labs(y = "Electricity Generated (Thousand Mwh)",
       x = "Year")

p.wrap2
```
</div>

***
<div style="text-align: center;">
### **Share and Growth of Wind and Solar Generation**
</div>

The generation of solar and wind power has experienced substantial growth between 2009 and 2023. Solar energy now accounts for nearly 5% of the total electricity generation, a remarkable increase from its previous share of just 0.03%. Additionally, wind power contributes approximately 14% to the total electricity generation, marking a significant rise from its previous 2.5% share.

<div style="text-align: center;">
```{r Share_and_Growth_of_Wind_Solar, echo=FALSE, message=FALSE, warning=FALSE}
share_growth <- df %>%
  dplyr::select(-Coal, -Natural_Gas, -Hydroelectric, -Nuclear) %>%
  dplyr::arrange(date) %>% 
  dplyr::mutate(Wind_Share = (Wind / Total_Generation) * 100, 
                Solar_Share = (Solar / Total_Generation) * 100) %>% 
  dplyr::mutate(Monthly_Solar_Change = c(NA, diff(Solar)),
                Monthly_Wind_Change = c(NA, diff(Wind)))

share <- share_growth %>% 
  ggplot(aes(x = date)) +
  geom_line(aes(y = Solar_Share, color = "Solar Share")) +
  geom_line(aes(y = Wind_Share, color = "Wind Share")) +
  scale_color_manual(values = c("Solar Share" = "yellow", "Wind Share" = "blue")) +
  theme(panel.background = element_rect(fill = "gray")) +
  labs(title = "Share of Wind and Solar Generation", 
       y = "Share of Total Generated (%)",
       x = "Year") +
  scale_x_date(date_labels = "%Y", date_breaks = "3 years")

change <- long %>% 
  dplyr::mutate(Month = month(date)) %>% 
  dplyr::mutate(Year = year(date)) %>% 
  dplyr::filter(Sources %in% c("Coal", "Solar", "Wind", "Natural_Gas", "Month", "Year")) %>% 
  dplyr::group_by(Month, Sources, Year) %>% 
  dplyr::summarise(Monthly_Generation = sum(Values)) %>% 
  dplyr::filter(Month == 1) %>% 
  dplyr::filter(Year == 2009 | Year == 2023) %>% 
  pivot_wider(names_from = Year,
              values_from = Monthly_Generation) %>% 
  dplyr::rename(Year_2009 = "2009",
                Year_2023 = "2023") %>% 
  dplyr::mutate(Growth = (Year_2023 - Year_2009) / Year_2009) %>% 
  dplyr::mutate(Growth = sprintf("%.2f%%", Growth * 100)) %>% 
  dplyr::ungroup(Month) %>% 
  dplyr::select(-Month) %>% 
  kbl(caption = "Growth of Generation Sources",
      col.names = c("Sources", "2009-01", "2023-01", "Growth")) %>% 
  kable_classic(full_width = F, html_font = "Cambria") %>% 
  row_spec(2, color = "red") %>% 
  row_spec(3, color = "green") %>% 
  row_spec(4, color = "blue")

change
plotly::ggplotly(share)

```
</div>

***
<div style="text-align: center;">
### **Periodic Patterns in Total, Wind, and Solar Generation**

#### **Total Generation Patterns**
</div>

The most influential factor affecting electricity generation is its seasonality, and there are four pivotal moments in the total generation where the trend undergoes changes:

1. Summer Peak: Demand peaks during the summer as longer daylight hours and need for air conditioning means more energy use
2. Fall Decrease: Temperatures begin to drop, reducing need for air conditioning. Shorter daylight hours means less energy use
3. Winter Increase: Low temperatures result in need for heating as well as the holiday season increases decorative lighting and energy use.
4. Spring Low: Temperatures begin to rise again reducing need for heating but not enough to warrant air conditioning, leading to lowest demand for energy.

We can also see the effect of the COVID-19 pandemic on generation as it hit an all time low in 2020. This historic decline was a consequence of the pandemic's ripple effects, including an economic deceleration, diminished commercial operations, constraints on travel, and disruptions in supply chains.

<div style="text-align: center;">
```{r Total_Generation_Graph, echo=FALSE, message=FALSE, warning=FALSE}
total <- df %>% 
  dplyr::select(date, Total_Generation) %>% 
  ggplot(aes(x = date,
             y = Total_Generation)) +
  geom_line() +
  labs(title = "Total Generation in the US (Monthly)",
       x = "Year",
       y = "Electricity Generated (Thousand Mwh)") +
  scale_x_date(date_labels = "%Y", date_breaks = "3 years")

plotly::ggplotly(total)
```
</div>

<div style="text-align: center;">
#### **Solar and Wind Generation Patterns**
</div>

Solar generation is directly affected by the amount of sunlight available. As a result, peak generation occurs between May - August (longer days, shorter nights) while lowest generation occurs between November - February (shorter days, longer nights).\

Wind generation is directly affected by the wind patterns. Wind generation climbs during the fall and peaks during the end of winter/beginning of spring where winds are the most consistent. Lowest generation occurs during the summer season when winds are the least consistent.\

There is a negative correlation of generation between solar and wind. Solar generation peaks during the summer while wind generation is at its lowest (vice versa).

<div style="text-align: center;">
```{r Solar_Wind_Generation, echo=FALSE, message=FALSE, warning=FALSE}
solar_wind <- long %>% 
  dplyr::filter(Sources %in% c("Solar", "Wind", "date")) %>% 
  ggplot(aes(x = date, 
             y = Values,
             col = Sources)) +
  geom_line() +
  facet_wrap(Sources ~ .,
             ncol = 1,
             scales = "free") +
  scale_x_date(date_labels = "%Y", date_breaks = "3 years") +
  labs(title = "Solar and Wind Generation",
       y = "Electricity",
       x = "Year")

plotly::ggplotly(solar_wind)
```
</div>
